Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-09-01 writeups/proposal_presentation/proposal_presentation_sv.qmd
Feedback: results, sbc, hmc

- Since then, more advanced algorithms such as Hamiltoninan Monte Carlo has enabled the estimation of richer and more complicated models

@@

$$
\begin{aligned}
y_t =& \space \beta exp(h_t/2) \epsilon_t \\
h_{t+1} =& \space \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
h_1 \sim& \space normal\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
\epsilon_t \sim &N(0,1) \space\space\space \eta_t \sim N(0,1)
\end{aligned}
$$

@@

## Kim Sherphard Chib (1998) Estimation strategy
:::{.nonincremental}
- Estimation of states using a Kalman filter, however, this requires the model to be linear and Gaussian.

@@

## Hamiltonian Monte Carlo (HMC)
- Developed in the physics literature (Duane et al. 1987) before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks (Nael, 1995) and statistical computing (Neal 2011).

- Key innovation: using the gradients of the target posterior distribution to generate an efficient path for the sampler to explore. It takes advantage of the geometry of the posterior to determine its next step.

- The Stan programming language's implementation of Hamiltonian Monte Carlo will be used for this study. Stan's default algorithm, the No-U-Turn Sampler (Hoffman & Gelman 2014), allows for direct sampling of the specified stochastic volatility model.

## Simulation Based Calibration (SBC)
<!-- Probably need to talk about what calibration means-->
<!-- - The procedure starts by taking draws from the prior distribution and creating datasets implied by each draw.

- Rank statistics are then calculated on the posterior samples estimated from the simulated data.

If B computation is correct it can capture the known parameers.-->
:::{.nonincremental}
SBC simulates data with known parameters and checks if on average, the interval of an estimated model contains the true parameter

1) Take a draw from the joint prior $\theta^{sim}$
2) Generate a dataset conditional on the prior draw
3) Calculate a rank statistics
4)  Compare the distribution of rank statistics to the discrete uniform distribution which arises when an algorithm is correctly calibrated.

@@

$$
\begin{aligned}
\pi(\theta) &= \int \pi(\theta|y^{sim}) \pi(y^{sim}|\theta^{sim}) \pi(\theta^{sim})dy^{sim} d\theta^{sim} 
\end{aligned}
$$
<!-- &= \int \pi(\theta|y^{sim}) \pi(y^{sim},\theta^{sim}) dy^{sim} d\theta^{sim}
 -->

@@

@@

- KSC's correction using importance sampling weights (MH correction step)
    + Reweights samples from approximate distribution such that reweighted draws come from the correct distribution

@@

>>>>

- Since then, more advanced algorithms such as Hamiltonian Monte Carlo has enabled the estimation of richer and more complicated models

@@

$$
\begin{aligned}
y_t =& \space \beta e^{h_t/2} \epsilon_t \\
h_{t+1} =& \space \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
h_1 \sim& \space normal\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
\epsilon_t \sim &N(0,1) \space\space\space \eta_t \sim N(0,1)
\end{aligned}
$$

@@

#  Estimation strategy

## Kim Sherphard Chib (1998)
:::{.nonincremental}
- Estimation of states using a Kalman filter, however, this requires the model to be linear and Gaussian.

@@

## Hamiltonian Monte Carlo (HMC)
- Key innovations:

    + Uses the gradients of the target posterior distribution to generate an efficient path for the sampler to explore.

    + Runs physics simulation - imagine the parameter is a marble rolling in random directions inside a bowl

    + Allows for direct sampling of the specified stochastic volatility model

- The Stan programming language's implementation of Hamiltonian Monte Carlo, the No-U-Turn Sampler (Hoffman & Gelman 2014), will be compared with KSC's strategy.

# Simulation Based Calibration (SBC)

@@

## Simulation Based Calibration (SBC)
The rank statistic for a given parameter and simulation follows a discrete uniform distribution (proven in Talts et al. (2018)).

@@

# Results

@@

- KSC's correction using importance sampling weights
    + Reweight samples from Gaussian mixture so that the new draws come from the correct distribution

@@

# Appendix

$$
\begin{aligned}
\pi(\theta) &= \int \pi(\theta|y^{sim}) \pi(y^{sim}|\theta^{sim}) \pi(\theta^{sim})dy^{sim} d\theta^{sim}  \\
&= \int \pi(\theta|y^{sim}) \pi(y^{sim},\theta^{sim}) dy^{sim} d\theta^{sim}
\end{aligned}
$$
